---
output:
  html_document: default
  pdf_document: default
---

# (PART) R Tutorials {-} 

# R Basics {#intro}

Before we begin, make sure you have [R](https://cran.r-project.org/) and 
[RStudio](https://www.rstudio.com/products/rstudio/download3/#download) installed.

## R Markdown

Throughout this course, [R Markdown](http://rmarkdown.rstudio.com/lesson-1.html) will make our lives easier. Make sure that the `rmarkdown` library is installed:

```
install.packages("rmarkdown")
```

For each assignment, you will create an R Markdown file (*.Rmd) and submit that file by the following class session using [classroom.google.com](https://classroom.google.com). Each class
has been made using R Markdown, so you can find many examples by going to the GitHub repository
for this course [github.com/jonpage/r-course](https://github.com/jonpage/r-course)

## Working with data already loaded into R

Base R comes with a set of sample data that is useful for illustrating techniques in R.
Run the following command to see a list of the datasets in the core library `datasets`:

```
library(help = "datasets")
```

These datasets are accessible automatically. We'll start with the Swiss Fertility and Socioeconomic Inicators (1888) dataset. See a description of the dataset by using the help command, either `?swiss` or `help(swiss)`. This dataset is technically a `data.frame`, which you can see by using the command `class(swiss)`. For more information on `data.frame`s take a look at the documentation(`help(data.frame)`)

### Numeric summaries

Here are a few ways we can summarize a dataset:

`head()` shows us the first six rows of a `data.frame`.
```{r into-head}
head(swiss)
```

`summary()` provides summary statistics for each column in a `data.frame`.
```{r intro-summary}
summary(swiss)
```

### Visual summaries

Scatterplot matrix (default plot of a data.frame):
```
plot(swiss)
# or
pairs(swiss)
```
```{r intro-plot, echo=FALSE}
plot(swiss)
```

Scatterplot of two dimensions
```
plot(swiss[,c("Education", "Fertility")])
# or
plot(swiss[4,1])
# or
plot(swiss$Education, swiss$Fertility)
# or
plot(swiss$Fertility ~ swiss$Education)
```
```{r intro-cross-plot, echo=FALSE}
plot(swiss$Fertility ~ swiss$Education)
```

Smoothed Scatterplot of two dimensions
```{r intro-smooth-scatter}
smoothScatter(swiss$Fertility ~ swiss$Examination)
```

Scatterplot with a `loess` (locally weighted polynomial regression)
```{r intro-smooth}
scatter.smooth(swiss$Fertility ~ swiss$Agriculture)
```

### Distribution plots

Histograms:
```{r intro-hist}
hist(swiss$Catholic)
```

Stem-and-Leaf Plots:
```{r intro-stem}
stem(swiss$Fertility)
```

Kernel density plot (and add a rug showing where observation occur):
```{r intro-rug}
plot(density(swiss$Fertility))
rug(swiss$Fertility)
```

Boxplots:
```{r intro-boxplot}
boxplot(swiss)
```


#### More complicated charts
Conditioning plots:
```{r intro-coplot}
coplot(swiss$Fertility ~ swiss$Examination | as.factor(swiss$Catholic > 50))
```

Star plots (half-star plots here):
```{r intro-star}
stars(swiss, key.loc = c(15,1), flip.labels = FALSE, full = FALSE)
```

## Assignment

Create a new R Markdown file. 

Choose a dataset from `datasets` (`library(help = "datasets")` will show you a list) and create 5 charts in an R Markdown file from the example charts above. Run the following command to see what else is available in the base R graphics package:

```
demo(graphics)
```
