---
output:
  html_document: default
  pdf_document: default
---

# (PART) R Tutorials {-} 

# R Basics {#intro}

Before we begin, make sure you have [R](https://cran.r-project.org/) and 
[RStudio](https://www.rstudio.com/products/rstudio/download3/#download) installed.

## R Markdown

Throughout this course, [R Markdown](http://rmarkdown.rstudio.com/lesson-1.html) will make our lives easier. Make sure that the `rmarkdown` library is installed:

```
install.packages("rmarkdown")
```

Intro to R and RStudio; Histograms, scatterplots, summary statistics

## Working with data already loaded into R

Base R comes with a set of sample data that is useful for illustrating techniques in R.
Run the following command to see a list of the datasets in the core library `datasets`:

```{r include=FALSE}
library(help = "datasets")
```

These datasets are accessible automatically. We'll start with the Swiss Fertility and Socioeconomic Inicators (1888) dataset. See a description of the dataset by using the help command, either `?swiss` or `help(swiss)`. This dataset is technically a `data.frame`, which you can see by using the command `class(swiss)`. For more information on `data.frame`s take a look at the documentation(`help(data.frame)`)

### Numeric summaries

Here are a few ways we can summarize a dataset:

`head()` shows us the first six rows of a `data.frame`.
```{r}
head(swiss)
```

`summary()` provides summary statistics for each column in a `data.frame`.
```{r}
summary(swiss)
```

### Visual summaries

```{r}
```

## Other stuff

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

