# Reading data {#read-data}

The first step in analyzing data with R is reading data into it. This lesson focuses on reading data, manipulating it with `dplyr` and a few summary visualizations.

## Data Source:

The US Census Bureau has a large selection of data on the population of the United States. 
The public-use micro surveys (PUMS) are available from the following link:

https://www.census.gov/programs-surveys/acs/data/pums.html

We'll take a look at the 1-year American Community Survey results for the state of Hawaii.

[Hawaii Population Records](https://www2.census.gov/programs-surveys/acs/data/pums/2015/1-Year/csv_phi.zip)

[The data dictionary](https://www2.census.gov/programs-surveys/acs/tech_docs/pums/data_dict/PUMSDataDict15.txt)

The specific file we are working with is the person record, so not every variable in the data dictionary will be available. Only those under the heading "PERSON RECORD" will be in the `csv_phi.zip` file.

## `read_csv`

To read in the downloaded file, we'll use the `readr` package, which you can install by installing `tidyverse`.

```{r read-data}
library(tidyverse)
pop_hi <- read_csv("data/csv_phi.zip")
pop_hi
```


## `dplyr`

Using the data dictionary we can identify some interesting variables.

```
PERSON RECORD

RT		1	
	Record Type		
		P .Person Record

AGEP		2	
	Age
		00 	.Under 1 year	
		01..99  .1 to 99 years (Top-coded***)

COW		1	
	Class of worker
		b .N/A (less than 16 years old/NILF who last
		  .worked more than 5 years ago or never worked)
		1 .Employee of a private for-profit company or
		  .business, or of an individual, for wages,
		  .salary, or commissions
		2 .Employee of a private not-for-profit,
		  .tax-exempt, or charitable organization
		3 .Local government employee (city, county, etc.)
		4 .State government employee
		5 .Federal government employee
		6 .Self-employed in own not incorporated
		  .business, professional practice, or farm
		7 .Self-employed in own incorporated
		  .business, professional practice or farm
		8 .Working without pay in family business or farm
		9 .Unemployed and last worked 5 years ago or earlier or never
                  .worked
		
SCHL		2	
	Educational attainment
		bb .N/A (less than 3 years old)
		01 .No schooling completed
		02 .Nursery school, preschool   
		03 .Kindergarten
		04 .Grade 1
		05 .Grade 2
		06 .Grade 3		
		07 .Grade 4
		08 .Grade 5
		09 .Grade 6
		10 .Grade 7		
		11 .Grade 8  
		12 .Grade 9
		13 .Grade 10
		14 .Grade 11		
		15 .12th grade - no diploma   
		16 .Regular high school diploma
		17 .GED or alternative credential
		18 .Some college, but less than 1 year
		19 .1 or more years of college credit, no degree
		20 .Associate's degree		
		21 .Bachelor's degree
		22 .Master's degree
		23 .Professional degree beyond a bachelor's degree
		24 .Doctorate degree
		
WAGP		6	
	Wages or salary income past 12 months
		bbbbbb 	 	.N/A (less than 15 years old)
		000000 		.None
		000001..999999 	.$1 to 999999 (Rounded and top-coded)

Note: Use ADJINC to adjust WAGP to constant dollars.


WKHP		2	
	Usual hours worked per week past 12 months
		bb 	 .N/A (less than 16 years old/did not work 
			 .during the past 12 months)
		01..98   .1 to 98 usual hours
		99 	 .99 or more usual hours
		
WKW		1	
	Weeks worked during past 12 months
		b .N/A (less than 16 years old/did not work 
		  .during the past 12 months)
		1 .50 to 52 weeks worked during past 12 months
		2 .48 to 49 weeks worked during past 12 months
		3 .40 to 47 weeks worked during past 12 months
		4 .27 to 39 weeks worked during past 12 months
		5 .14 to 26 weeks worked during past 12 months 
		6 .less than 14 weeks worked during past 12 months
		
ESR		1	
	Employment status recode
		b .N/A (less than 16 years old)
		1 .Civilian employed, at work
		2 .Civilian employed, with a job but not at work
		3 .Unemployed
		4 .Armed forces, at work
		5 .Armed forces, with a job but not at work
		6 .Not in labor force
		
PERNP		7	
	Total person's earnings
		bbbbbbb 		.N/A (less than 15 years old)
		0000000 		.No earnings
		-010000 		.Loss of $10000 or more (Rounded & bottom-coded 
       					.components) 
		-000001..-009999 	.Loss $1 to $9999 (Rounded components)
		0000001 		.$1 or break even
		0000002..9999999 	.$2 to $9999999 (Rounded & top-coded components)
		
Note: Use ADJINC to adjust PERNP to constant dollars.

PINCP		7	
	Total person's income (signed)
		bbbbbbb 		.N/A (less than 15 years old)
		0000000 		.None
		-019999 		.Loss of $19999 or more (Rounded & bottom-coded 
       					.components) 
		-000001..-019998 	.Loss $1 to $19998 (Rounded components)
		0000001 		.$1 or break even
		0000002..9999999 	.$2 to $9999999 (Rounded & top-coded components)
			
Note: Use ADJINC to adjust PINCP to constant dollars.
```

Let's focus on employed civilians (ESR either 1 or 2) working full time (WKHP > 32) for close to the entire year (WKW either 1 or 2).
```{r read-filter, dependson="read-data"}
pop_hi <- pop_hi %>%
  filter(
    ESR %in% c(1, 2),
    as.numeric(WKHP) > 32,
    WKW %in% c(1, 2)
  )
```

If you are unsure if a column that you want to treat as numeric contains letters, you can run the following
command to get a list of the values containing letters:
```{r read-alpha, dependson="read-filter"}
grep("[[:alpha:]]", pop_hi$WKHP, value = TRUE)
```

We can use two functions to add new columns (or change existing ones).

1. `mutate()` adds columns and keeps the previous columns
2. `transmute()` adds columns and removes the previous columns

This time we want to drop the columns we don't mention.
```{r read-factor, dependson="read-filter"}
pop_hi <- pop_hi %>%
  transmute(
    age = as.numeric(AGEP),
    worker_class = factor(COW, labels = c(
      "for-profit",
      "not-for-profit",
      "local government",
      "state government",
      "federal government",
      "self-employed not incorporated",
      "self-employed incorporated",
      "family business no pay"
    )),
    school = SCHL,
    wages = as.numeric(WAGP),
    top_coded_wages = WAGP == 999999
)
```

Creating a custom factor variable for educational attainment: 
```{r read-education, dependson="read-factor"}
education_levels <- c("less than HS", "HS", "associates", "bachelors", "masters", "doctorate")

pop_hi$education <- NA
pop_hi[pop_hi$school < 16,]$education <- "less than HS"
pop_hi[pop_hi$school > 16 & pop_hi$school < 20,]$education <- "HS"
pop_hi[pop_hi$school == 20,]$education <- "associates"
pop_hi[pop_hi$school == 21,]$education <- "bachelors"
pop_hi[pop_hi$school %in% c(22, 23),]$education <- "masters"
pop_hi[pop_hi$school == 24,]$education <- "doctorate"

pop_hi <- pop_hi %>%
  mutate(education = factor(education, levels = education_levels))
```

## First Look at `ggplot2`

See the [ggplot2 documentation](ggplot2.tidyverse.org/reference/) for details and inspiration.

```{r read-point, dependson="read-education"}
ggplot(pop_hi, aes(age, wages)) +
  geom_point()
```

Income has a skewed distribution, so it is often presented/analyzed in logs. Here's how to modify the above 
chart to display income in logs:

```{r read-scale, dependson="read-education"}
pop_hi <- pop_hi %>%
  mutate(log_safe_wages = ifelse(wages == 0, 1, wages))
pop_hi %>%
  ggplot(aes(age, log_safe_wages)) +
  geom_point() +
  scale_y_log10()
```

## Heatmaps

Heatmaps allow you to get a sense of the concetration of observations in regions where there are many
overlapping points:

```{r read-bin2d, dependson="read-scale"}
ggplot(pop_hi, aes(age, log_safe_wages)) +
  geom_bin2d() +
  scale_y_log10()
```

## Hexbins

Hexbins use hexagons instead of squares, which helps avoid the rectangular sections in
heatmaps that may misrepresent your data.

You will need to install the `hexbin` package.
```
install.packages("hexbin")
```

```{r read-hex, dependson="read-scale"}
pop_hi %>%
  filter(wages > 10000, wages < 300000) %>%
  ggplot(aes(age, wages)) +
  geom_hex() + 
  scale_x_log10()
```

Here we can see that inequality in wages for workers increases with age.

## Other topics from this dataset

This dataset includes information on the majors for degree holders and the industry codes. You could use that
additional information to ask how well targeted majors are to particular industries and how incomes vary
across choice of major. Because of the size of the data in the Hawaii sample, it would be better to ask some
of these questions at the national level.

Go to the [ACS PUMS documentation page](https://www.census.gov/programs-surveys/acs/technical-documentation/pums/documentation.2015.html) for more information.


## Assignment

Choose another pair of variables from the data dictionary and visualize them with a scatterplot,
a heatmap, and a hexbin plot.
