---
output:
  html_document: default
  pdf_document: default
---

# Lines and Curves {#lines}

## Data

We will use data on the number of active duty personnel in Hawaii. 
The first dataset is an Excel file pulled from the State of Hawaii Department of Business, Economic Development, and
Tourism (DBEDT) [2015 State of Hawaii Data Book](http://dbedt.hawaii.gov/economic/databook/2015-individual/). See the line listed
as, "10.03 - Active Duty Personnel, by Service: 1953 to 2015."
The data is originally from the [US Defense Manpower Data Center](www.dmdc.osd.mil/appj/dwp/stats_reports.jsp)

```{r}
library(tidyverse)
library(readxl)
```

```{r, echo=FALSE}
select <- dplyr::select
```

```{r}
mil_personnel <- read_excel("data/100315.xls", range = "A5:L38", col_types = "numeric")
mil_personnel <- bind_rows(
  mil_personnel %>% select(1:6) %>% magrittr::set_colnames(c("Year", "Total", "Army", "Navy", "Marine Corps", "Air Force")),
  mil_personnel %>% select(7:12) %>% magrittr::set_colnames(c("Year", "Total", "Army", "Navy", "Marine Corps", "Air Force"))
)
mil_personnel
```

Notice that the `Year` 2015 was turned into `NA`. This happened because the value in the corresponding cell was '2/ 2015'. Let's remove the final row of `NA`s and replace the remaining `NA` with 2015. 

```{r}
mil_personnel <- mil_personnel %>% filter(!is.na(Total))
mil_personnel[is.na(mil_personnel$Year),]$Year <- 2015
mil_personnel
```


## geom_smooth

`geom_smooth` allows you to have smooth lines appear in your chart. With no argument, it will
choose `loess` for series shorter than 1,000 observations. It shows a shaded confidence interval.

```{r}
mil_personnel %>%
  ggplot(aes(Year, Total)) +
  geom_point() +
  geom_smooth()
```

Here's what it looks like if we fit a linear model instead:

```{r}
mil_personnel %>%
  ggplot(aes(Year, Total)) +
  geom_point() +
  geom_smooth(method = "lm")
```

We can also just have a line chart that connects the points:

```{r}
mil_personnel %>%
  ggplot(aes(Year, Total)) +
  geom_point() +
  geom_line()
```

## geom_abline

`geom_abline` allows you to display lines with a specific intercept and slope. If no intercept or slope is
provided, a 45-degree line will be shown.

```{r}
x = rnorm(100)
y = 2.5 + 1.2 * x + rnorm(100)
test_data <- data_frame(x, y)

test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() 
```

```{r}
test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() +
  geom_abline(intercept = 2.5, slope = 1.2, color = "red") 
```

## geom_vline

`geom_vline` allows you to draw vertical lines by specifying an x intercept.

```{r}
test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() +
  geom_abline(intercept = 2.5, slope = 1.2, color = "red") +
  geom_vline(xintercept = 2, color = "blue")
```

## hline

`geom_vline` allows you to draw vertical lines by specifying an x intercept.

```{r lines-vline}
test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() +
  geom_abline(intercept = 2.5, slope = 1.2, color = "red") +
  geom_vline(xintercept = 2, color = "blue") +
  geom_hline(yintercept = 1, color = "#4FCC53", lty = 2)
```

## Assignment

Create a visualization of the military data by branch (i.e., `Army`, `Navy`, etc.) using `facet_wrap()`. Plot both the points and a smooth line.

The data we have been working with is not yet [tidy](http://tidyr.tidyverse.org/). Each row contains 
multiple observations (observations for Army, Navy, etc.). To make this tidy we should have one column with 
the personnel counts and one column that indicates the branch.

```{r}
tidy_mil <- mil_personnel %>%
  gather(branch, personnel, -Year)
tidy_mil
```
