---
output:
  html_document: default
  pdf_document: default
---

# Lines and Curves

## Data

We will use data on the number of active duty personnel in Hawaii. 
The first dataset is an Excel file pulled from the State of Hawaii Department of Business, Economic Development, and
Tourism (DBEDT) [2015 State of Hawaii Data Book](http://dbedt.hawaii.gov/economic/databook/2015-individual/). See the line listed
as, "10.03 - Active Duty Personnel, by Service: 1953 to 2015."
The data is originally from the [US Defense Manpower Data Center](www.dmdc.osd.mil/appj/dwp/stats_reports.jsp)

```{r}
library(readxl)
library(dplyr)
library(magrittr)
mil_personnel <- read_excel("data/100315.xls", range = "A5:L38", col_types = "numeric")
mil_personnel <- bind_rows(
  select(mil_personnel, 1:6) %>% set_colnames(c("Year", "Total", "Army", "Navy", "Marine Corps", "Air Force")),
  select(mil_personnel, 7:12) %>% set_colnames(c("Year", "Total", "Army", "Navy", "Marine Corps", "Air Force"))
) %>%
  filter(!is.na(Year))
mil_personnel
```


## geom_smooth

`geom_smooth` allows you to have smooth lines appear in your chart. With no argument, it will
choose `loess` for series shorter than 1,000 observations. It shows a shaded confidence interval.

```{r}
library(ggplot2)
mil_personnel %>%
  ggplot(aes(Year, Total)) +
  geom_point() +
  geom_smooth()
```

Here's what it looks like if we fit a linear model instead:

```{r}
mil_personnel %>%
  ggplot(aes(Year, Total)) +
  geom_point() +
  geom_smooth(method = "lm")
```

We can also just have a line chart that connects the points:

```{r}
mil_personnel %>%
  ggplot(aes(Year, Total)) +
  geom_point() +
  geom_line()
```

## geom_abline

`geom_abline` allows you to display lines with a specific intercept and slope. If no intercept or slope is
provided, a 45-degree line will be shown.

```{r}
x = rnorm(100)
y = 2.5 + 1.2 * x + rnorm(100)
test_data <- data_frame(x, y)

test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() 
```

```{r}
test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() +
  geom_abline(intercept = 2.5, slope = 1.2, color = "red") 
```

## geom_vline

`geom_vline` allows you to draw vertical lines by specifying an x intercept.

```{r}
test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() +
  geom_abline(intercept = 2.5, slope = 1.2, color = "red") +
  geom_vline(xintercept = 2, color = "blue")
```

## hline

`geom_vline` allows you to draw vertical lines by specifying an x intercept.

```{r}
test_data %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  xlim(-2, 6) + ylim(-2, 6) +
  coord_fixed() +
  geom_abline() +
  geom_abline(intercept = 2.5, slope = 1.2, color = "red") +
  geom_vline(xintercept = 2, color = "blue") +
  geom_hline(yintercept = 1, color = "#4FCC53", lty = 2)
```
