---
output:
  html_document: default
  pdf_document: default
---

# ggplot2 Extensions and Scatterplot Matrices (GGally) {#ggplot-exts}

## Data 

For this section, we'll look at data from the American Community Survey (ACS)
on immigration. To download the data,

1. Go to the [American FactFinder website](https://factfinder.census.gov/).
2. Click on the "Download Center" section, then click the "DOWNLOAD CENTER" button.
3. Click the "NEXT" button, since we know the table we want to download.
4. Select "American Community Survey" from the Program dropdown.
5. Select "2015 ACS 5-year estimates", click the "ADD TO YOUR SELECTIONS" button, then click "NEXT"
6. Select "County - 050" from the geographic type dropdown, then select "All Counties within United States", click the "ADD TO YOUR SELECTIONS" button, then click "NEXT"
7. Type `income mobility` in the "topic or table name"" search box, then select the option that reads: 

"B07011: MEDIAN INCOME IN THE PAST 12 MONTHS (IN 2015 INFLATION-ADJUSTED DOLLARS) BY GEOGRAPHICAL MOBILITY IN THE PAST YEAR FOR CURRENT RESIDENCE IN THE UNITED STATES"

Click "GO", then check the checkbox beside the table we found. Now click on the "Download" button, and 
uncheck the option that says, "Include descriptive data element names." Click "Ok" to create your zip file. Once the file has been created, click "DOWNLOAD" to download the zip file.

The following will assume you moved the following files within the zip to your `data` folder:

* ACS_15_5YR_B07011_with_ann.csv
* ACS_15_5YR_B07011_metadata.csv

The file `ACS_15_5YR_B07011.txt` tells us how to interpret codes within our data. It is possible for median values to be followed by a `+` or `-` if they
are in the upper or lower open-ended interval. In our dataset we don't have any medians in the upper open-ended interval, but we do have entries in the
lower open-ended interval.

```{r}
library(readr)
acs <- read_csv("data/ACS_15_5YR_B07011_with_ann.csv", col_types = strrep("c", 15), na = c("-", "(X)"))
meta <- read_csv("data/ACS_15_5YR_B07011_metadata.csv")
meta
```


Let's keep only the variables we care about, using more informative variable names.
```{r}
library(dplyr)
acs_mobility <- acs %>%
  transmute(
    geo = `GEO.display-label`,
    same_house = HD01_VD03,
    same_county = HD01_VD04,
    same_state = HD01_VD05,
    same_country = HD01_VD06,
    different_country = HD01_VD07
  )
acs_mobility
```

Now we can add an indicator for whether the median value is in the lowest available interval. This would mean that the median value presented has been bottom-coded.
```{r}
acs_mobility <- acs_mobility %>%
  mutate(
    same_country_bc = grepl("[0-9]*-", same_country),
    different_country_bc = grepl("[0-9]*-", different_country)
  )
acs_mobility
```

Let's see how many counties have observations that are bottom coded:
```{r}
acs_mobility %>%
  summarize(same_country_bc = sum(same_country_bc), different_country_bc = sum(different_country_bc), counties = n())
```

Let's see what the typical bottom-coded values are:
```{r}
acs_mobility %>%
  summarize(sc_median = median(same_country[same_country_bc]))
```

Since this is a small number of counties given the entire range, we'll 

## ggplot2 extensions

There are many extensions the community have made that build on ggplot2. 
The following link provides a gallery of many of these extensions: 

[ggplot2 extensions](http://www.ggplot2-exts.org/gallery)

## scatterplot matrix (GGally::scatmat)

One particular library, [GGally](http://ggobi.github.io/ggally/), has a great set of visualizations
to extend those that come prebuilt with `ggplot`. One common visualization tool that is missing from
ggplot is the scatterplot matrix. While base R provides `splom()` in the `lattice` library, 
`GGally::ggpairs` and `GGally::ggscatmat` provide an easy tool to create a scatterplot matrix with
ggplot2.


## GGally::glyphs
