# Log scales {#log}

The focus of this lecture is on log scales. The goal is to give you an
intuition about when to use log scales and how to interpret them. We will
refresh on basic log transformations and see how they affect how we can encode
data in a variety of visualizations.

This lecture uses the following packages:
```
tidyverse
```

## Basic Log Review

### Order of magnitude

Remember that the log of a number is the exponent required to transform the base into the
input.

So that,

\[ log_{10} (x) = y  \]

implies

\[ 10^y = x \]

Let's use a concrete example where $x$ is a vector of the integers from 1 to 100.

```{r}
x <- 1:100
x
```

Let's store the base 10 log of $x$ in the vector $y$:

```{r}
y <- log(x, base = 10)
y <- log10(x)
y
```

Notice in particular the following pattern:

```{r}
log10(c(1, 10, 100, 1000, 10000))
```

It is useful to think of the logarithm (log) as recording the order of magnitude of the input value.

```{r}
library(tidyverse)
ggplot(data.frame(x = x, y = y), aes(x, y)) +
  geom_point()
```

From looking at the scatter plot of $x$ and $y$ you should notice that the space between larger
values is diminished relative to the space between smaller values.

### Percent change

A useful feature of logged variables is that the difference between two logged values is proportional to
the percentage change in the original values. So, if the difference between two logged values is the same
as the difference between two other logged values, the percentage change across both sets is the same.

```{r}
log10(110) - log10(100)
(110 - 100)/100
log10(220) - log10(200)
(220 - 200)/200
```

This is useful in visualizations because similar gaps in a log variable in different regions of the
chart correspond to the same percentage change.

To test this let's create a variable that grows by a fixed percentage.

```{r}
constant_growth = 100 * (1.10) ^ (1:100)
constant_growth
```

```{r}
qplot(1:100, constant_growth)
```

```{r}
qplot(1:100, log(constant_growth))
```

`diff(x)` returns a vector of the differences between consecutive values of `x`:

```{r}
diff(constant_growth)
```

Taking a look at the differences of the logged constant growth variable we see that
we the change across values is now constant.

```{r}
diff(log(constant_growth))
```


### Histogram comparison

We can also compare the histogram of a normal distribution to a distribition that
would be normal if we applied logs.
```{r}
qplot(rnorm(10000, mean = 10, sd = 1), main = "Normal Distribution")
```

```{r}
qplot(exp(rnorm(10000, mean = 10, sd = 1)), main = "Log-normal distribution")
```


## Data

For a practical application of logs we'll look back at income. 
This time we're using zip-code level data from the 
[IRS Statistics of Income (SOI)](https://www.irs.gov/statistics/soi-tax-stats-statistics-of-income)
program.

There are a variety of datasets available on [individual income (form 1040)]( https://www.irs.gov/statistics/soi-tax-stats-individual-income-tax-return-form-1040-statistics).
From the main [zip-code data page](https://www.irs.gov/statistics/soi-tax-stats-individual-income-tax-statistics-zip-code-data-soi), click the link to the [2015 data](https://www.irs.gov/statistics/soi-tax-stats-individual-income-tax-statistics-2015-zip-code-data-soi), then download the 
[CSV for all states without AGI](https://www.irs.gov/pub/irs-soi/15zpallnoagi.csv). The file
with AGI breaks down the observations further into ranges of adjusted gross income (AGI).
We will instead focus on the zip-level data across all AGI groups.
You will also want to [download the documentation](https://www.irs.gov/pub/irs-soi/15zpdoc.doc) 
to determine which variables we want to  keep and how to interpret them.

```{r}
raw_irs <- read_csv("data/15zpallnoagi.csv")

irs <- raw_irs %>%
  filter(ZIPCODE != '00000') %>% # remove state-level summary
  transmute(
    zip = ZIPCODE,
    state = STATE,
    households = N1,
    population = N2,
    agi = A00100,
    agi_pc = agi / population,
    total_income = A02650,
    wages = A00200,
    farms = SCHF,
    farm_proportion = farms / households,
    taxes = A10300,
    taxes_pc = taxes / population,
    taxes_agi = taxes / agi,
    taxes_total_income = taxes / total_income
  )
as.tibble(irs)
```


## Logs in position

### Population size and AGI per capita

Let's take a look first at the distribution of population and 
adjusted gross income per capita across zipcodes:

```{r}
ggplot(irs, aes(population, agi_pc)) +
  geom_point()
```

Remember that the log transformation compresses the space between larger values. Our scatter plot
indicates that smaller values are currently more compressed. This is a signal that we should log transform
our variables. The log transformation is useful for some of the assumptions of linear modelling. 
In the visual explorations here, the choice to log transform is largely aesthetic. If log transforming
a variable makes it easier to visually inspect and understand your data, than it is useful.

### Histogram comparison

Let's look at the histogram for the `population`:
```{r}
ggplot(irs, aes(population)) + geom_histogram()
```

And now the logged version:
```{r}
ggplot(irs, aes(population)) + geom_histogram() + scale_x_log10()
```

While the logged version of `population` feels closer to a normal distribution, it is clearly not
normally distributed. Even though population across zip codes is not exactly log-normally distributed,
it is still useful for creating easier to interpret and analyze visualizations.

```{r}
ggplot(irs, aes(population, agi_pc)) +
  geom_point(alpha = 0.1) +
  geom_smooth() +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10()
```


### Taxes and Farms

Let's plot two proportions: the proportion of returns in a zipcode representing farms, `farm_proportion`,
and the proportion of total income that is tax liability, `taxes_total_income`.

```{r}
ggplot(irs, aes(farm_proportion, taxes_total_income)) +
  geom_point() + 
  geom_smooth()
```

While the data is densely packed near the origin, it does not mean it would be appropriate to
log these variables. Keep in mind that the log of 0 is undefined and in the limit 
from the right is negative infinty. We can manually remove these values with a filter.

```{r}
ggplot(irs, aes(farm_proportion, taxes_total_income)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(labels = scales::percent) +
  scale_y_log10(labels = scales::percent) 
```


## Logs in color

Let's tweak the last visualization by encoding `agi_pc` in the color of the points.

```{r}
ggplot(irs %>% filter(farm_proportion > 0), aes(farm_proportion, taxes_total_income, color = agi_pc)) +
  geom_point() +
  scale_x_log10(labels = scales::percent) +
  scale_y_log10(labels = scales::percent) 
```

Using the transformation `log10` (see `trans` in the `?continuous_scale` documentation), we
get a more gradual shift in color that makes it easier to see the changes in per capita
adjusted gross income.


```{r}
ggplot(irs %>% filter(farm_proportion > 0), aes(farm_proportion, taxes_total_income, color = agi_pc)) +
  geom_point() +
  scale_x_log10(labels = scales::percent) +
  scale_y_log10(labels = scales::percent) +
  scale_color_continuous(trans = "log10")
```

We can finally change the `low` and `high` colors and add `alpha` to make our scatter plot easier to read.
Note that I have added a log transformation to the size, which is tied to the `population` variable. Remove
the log transformation to see the difference.

```{r}
library(scales)
ggplot(irs %>% filter(farm_proportion > 0), aes(farm_proportion, taxes_total_income, color = agi_pc)) +
  geom_point(aes(size = population), alpha = 0.1) +
  scale_x_log10(labels = scales::percent) +
  scale_y_log10(labels = scales::percent) +
  scale_color_continuous(trans = "log10", low = scales::muted("red"), high = scales::muted("blue")) +
  scale_size_continuous(trans = "log10")
```


## Assignment

Choose two different variables from the `irs` dataset to visualize. Choose whether or not to log each variable
in your visualization and explain why that was the right choice. Show at least one alternative (logged version
vs. raw values) visualization and discuss how it compares to your preferred choice.
