# Time-Series Modeling {#time-series}

This lecture uses the following packages:

```
tidyverse
tidyquant
vars
```

## Introduction

The focus of this lecture is on time-series data. We will be making use of a
new package that helps us apply the tools of the tidyverse to time series. To
read more about the `tidyquant` package, check out its website:

https://business-science.github.io/tidyquant/

## Data

To explore time-series modeling, we will download a few macroeconomic time series from  
[FRED](https://fred.stlouisfed.org/). The data list I compiled for this lecture can be
accessed at the following link:

https://research.stlouisfed.org/pdl/988

The following steps assume you used the `Zipped Tab Delimted Text` option with `1972-01-01`
as the start date.

```{r}
library(tidyverse)

daily <- read_tsv(
  "data/Time_series_lecture_txt/Time_series_lecture_Daily.txt",
  na = c(".")
)
monthly <- read_tsv(
  "data/Time_series_lecture_txt/Time_series_lecture_Monthly.txt",
  na = c(".")
)
quarterly <- read_tsv(
  "data/Time_series_lecture_txt/Time_series_lecture_Quarterly.txt",
  na = c(".")
) %>%
  mutate(GDPC1_CHG = c(NA, diff(GDPC1)))
```

Since we have data in multiple frequencies, we first need to aggregate up to the
quarterly level.

```{r}
library(tidyquant)
daily_q <- daily %>%
  tq_transmute(mutate_fun = to.quarterly)
monthly_q <- monthly %>%
  tq_transmute(mutate_fun = to.quarterly)
all_q <- quarterly %>%
  mutate(DATE = as.yearqtr(DATE, format = "%Y-%m-%d")) %>%
  merge(monthly_q, all = TRUE) %>%
  merge(daily_q, all = TRUE)
head(all_q)
```


## Hold-Out Set

Just like in the previous lecture we want to use a training set so that
we can evaluate the accuracy of our model on new data. In the context of
time-series data, the validation/test sets are usually referred to as
the hold-out set.

```{r}
train <- all_q %>% filter(DATE < "2010 Q1")
hold_out <- all_q %>% filter(DATE >= "2010 Q1")
```

## GDP

Let's begin by plotting the series we want to predict (GDP):
```{r}
ggplot(train, aes(x = DATE, y = GDPC1)) +
  geom_line() +
  scale_x_yearqtr() +
  labs(title = "Real GDP")
```

The GDP series in [not stationary](https://en.wikipedia.org/wiki/Stationary_process) 
(you can see that the mean changes over time). Let's look at `GDPC1_CHG`, which is real 
GDP change from one quarter to the next.

```{r}
ggplot(train) +
  geom_line(aes(DATE, GDPC1_CHG)) +
  scale_x_yearqtr() +
  labs(title = "Real GDP Change")
```

This series appears stationary so we'll use it when the model we look at requires a stationary series.

## Autoregressive Model

The Autoregressive (AR) model says that previous values are our best
predictors of the future. 

Here is the equation for an AR(1) model:

\[ y_t = \rho y_{t-1} + \varepsilon_t \]

$y_{t-1}$ is the value last period.

There are many options possible with the `ar()` function, but we will
stick to the defaults.

```{r}
ar_model <- ar(train$GDPC1_CHG, na.action = na.omit)
ar_model
```

### AR Performance

```{r}
ar_prediction <- predict(ar_model, newdata = c(0), n.ahead = 12)
ar_prediction
```

```{r}
ggplot(cbind(hold_out[1:12,c("DATE", "GDPC1_CHG")], as.data.frame(ar_prediction)), aes(x = DATE)) +
  geom_ribbon(aes(ymin = pred - se, ymax = pred + se), alpha = 0.25, fill = scales::muted("green")) +
  geom_line(aes(y = pred), lty = 2) +
  geom_line(aes(y = GDPC1_CHG)) +
  scale_x_yearqtr() +
  scale_y_continuous() +
  labs(title = "AR prediction of GDP Change", subtitle = "Actual = solid, prediciton = dashed, se = green")
```

## ARIMA

An Autoregressive integrated moving average (ARIMA) model is able to model non-stationary series.
Just like the `ar()` function, the `arima()` function has many options for tuning the 
results. Again we will stick to the defaults, but we do need to specify the `order` parameter. 
The order is a three integer vector, (`p`, `d`, `q`), where `p` is the autoregressive
order (above we had 2), `d` is the degree of differecing (we implicitly assumed this to be 1), 
and `q` is the moving average order. 
Since ARIMA can handle non-stationary series we will model `GDPC1`.

```{r}
arima_model <- arima(train$GDPC1, c(2, 1, 0))
arima_model
```

### ARIMA Performance

```{r}
arima_prediction <- predict(arima_model, n.ahead = 12)
arima_prediction
```


We can compare the ARIMA prediction to the actual values.

```{r}
ggplot(cbind(hold_out[1:12,c("DATE", "GDPC1")], as.data.frame(arima_prediction)), aes(x = DATE)) +
  geom_ribbon(aes(ymin = pred - se, ymax = pred + se), alpha = 0.25, fill = scales::muted("green")) +
  geom_line(aes(y = pred), lty = 2) +
  geom_line(aes(y = GDPC1)) +
  scale_x_yearqtr() +
  scale_y_continuous() +
  labs(title = "ARIMA prediction of GDP", subtitle = "Actual = solid, prediciton = dashed, se = green")
```


## Vector Autoregression

Another approach is using a system of variables and allowing old values of each variable
to affect the others.

```{r}
library(vars)
var_model <- VAR(train %>% dplyr::select(GDPC1, CBIC1, PERMIT), p = 2)
var_model
```

```{r}
var_prediction <- predict(var_model, n.ahead = 12)
var_prediction
```

```{r}
ggplot(
  cbind(hold_out[1:12,c("DATE", "GDPC1")], as.data.frame(var_prediction$fcst$GDPC1)), 
  aes(x = DATE)
) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25, fill = scales::muted("green")) +
  geom_line(aes(y = fcst), lty = 2) +
  geom_line(aes(y = GDPC1)) +
  scale_x_yearqtr() +
  scale_y_continuous() +
  labs(
    title = "VAR prediction of GDP", 
    subtitle = "Actual = solid, prediciton = dashed, 95% CI = green"
  )
```

## Assignment

Pick your preferred model and test its performance in the entire hold-out dataset.
Report the RMSE (see the [previous lesson](#cross-setion)) and create a chart like
the ones above showing lines for the actual and predicted values with a blue ribbon
indicating one standard error (for AR or ARIME) or the confidence interval (for VAR)
about the prediciton.
