---
output:
  html_document: default
  pdf_document: default
---

# (PART) Topics {-} 

# Anscombe's Quartet {#anscombe}

Anscombe quartet emphasizes the need to move beyond basic numerical summaries of your data.
The `anscombe` dataset has four sets of `x` and `y` variables with very similar summaries, 
but distinct visual patterns

## Prep the data

```{r}
anscombe
```

First we'll use `tidyr` to reshape the anscombe dataset to make it easier to work with.

```{r}
library(tidyr)
tidy_anscombe <- anscombe %>%
  mutate(id = row_number()) %>%
  gather(key = key, value = value, everything(), -id)
tidy_anscombe
```

```{r}
tidy_anscombe <- tidy_anscombe %>%
  separate(key, c("x_or_y", "series"), 1)
tidy_anscombe
```

```{r}
tidy_anscombe <- tidy_anscombe %>%
  spread(x_or_y, value)
tidy_anscombe
```


## Numeric summary

```{r}
tidy_anscombe %>%
  group_by(series) %>%
  summarise(
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x = sd(x),
    sd_y = sd(y),
    cor = cor(x, y)
  )
```



## Visual summary

While the visual summaries suggest very similar datasets, the visual summaries help identify the differences:

```{r}
library(ggplot2)
tidy_anscombe %>%
  ggplot(aes(x, y)) +
  geom_point() +
  facet_wrap(~ series) +
  coord_fixed()
```


## The Datasaurus Dozen

The Datasaurus Dozen is a set of series, like the Anscombe's quartet with similar numerical summaries and
radically different visual summaries. See a great discussion of this dataset by the creators, Justin Matejka and
George Fitzmaurice [here](https://www.autodeskresearch.com/publications/samestats)

Download the data [here](https://www.autodeskresearch.com/sites/default/files/The%20Datasaurus%20Dozen.zip) and 
move the DatasaurusDozen.tsv file into your data folder.

```{r}
datasaurus <- read_tsv("data/DatasaurusDozen.tsv")
datasaurus %>%
  group_by(dataset) %>%
  summarise(
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x = sd(x),
    sd_y = sd(y),
    cor = cor(x, y)
  )
```

Visual summaries
```{r}
datasaurus %>%
  ggplot(aes(x, y)) +
  geom_point() +
  facet_wrap(~ dataset, ncol = 6) +
  coord_fixed()
```
